---
title: "Grandparents Quant Project"
output: html_notebook
---


```{r, echo=FALSE}
##Packages
library(tidyverse)
library(knitr)
library(reshape2)
```

```{r, echo=FALSE, message=FALSE}
##Read in the data
master <- read_csv("Data/Master.csv")
```

In a previous life, I worked as an academic in Australia, and whilst there I completed my PhD on grandparent care.


##Table 1: Numbers and proportions of Australian children in care by type of care, 2011-2017

```{r, echo=FALSE}
filterwords <- c("Usually","usually","Total","Type", "Formal","Informal")
master %>%
  select(Care_Type, Total_Children_No_No_2017, Total_Children_Prop_No_2017, Total_no_no_2014, Total_no_prop_2014, Total_no_no_2011, Total_no_prop_2011) %>%
  filter(!grepl(paste(filterwords, collapse="|"), Care_Type)) %>%
  rename(`Total 2017 (000s)` = Total_Children_No_No_2017,
         `Proportion 2017` = Total_Children_Prop_No_2017,
         `Total 2014 (000s)` = Total_no_no_2014,
         `Proportion 2014` = Total_no_prop_2014,
         `Total 2011 (000s)` = Total_no_no_2011,
         `Proportion 2011` = Total_no_prop_2011,
         `Care Type` = Care_Type) %>%
  arrange(desc(`Total 2017 (000s)`)) %>%
  kable()
  
```

As seen above, in 2017 around 1 in 5 Australian children were cared for regularly by their grandparents. However, this was down from 1 in 4 children in 2011. In contrast, formal types of care, notably long day care and before and/or after school care, are both up since 2011.

Indeed, breaking down the data by type of care received, we can see that the proportion of Australian children in formal care is rising over time, whilst the proporiton of kids in informal care has fallen slightly. This has occured as the proportion of Australian children _not_ in any type of care has risen by XXX between 2011 and 2017.

```{r, echo=FALSE, dpi=600}
master %>%
  filter(grepl("Usually", Care_Type, ignore.case=TRUE)) %>%
  filter(Care_Type!="Usually attended care") %>%
  select(Care_Type, Total_no_prop_2011, Total_no_prop_2014, Total_Children_Prop_No_2017) %>%
  rename_at(vars(-Care_Type), function(x) gsub("Total_no_prop_", "", x)) %>%
  rename(`2017` = Total_Children_Prop_No_2017) %>%
  reshape2::melt(id.vars="Care_Type") %>%
  ggplot() +
  aes(variable, value, group=Care_Type, color=Care_Type) +
  geom_line() +
  theme_minimal() +
  ylab("% of Australian children") +
  xlab("Year") +
  theme(legend.title = element_blank())
```

There has, in fact, been a fall in all types of informal childcare:

```{r, echo=FALSE}
informalcare <- c("Non-resident parent","Grandparent","Other person","Other relative")
master %>%
  filter(Care_Type %in% informalcare) %>%
    select(Care_Type, Total_no_prop_2011, Total_no_prop_2014, Total_Children_Prop_No_2017) %>%
  rename_at(vars(-Care_Type), function(x) gsub("Total_no_prop_", "", x)) %>%
  rename(`2017` = Total_Children_Prop_No_2017) %>%
  reshape2::melt(id.vars="Care_Type") %>%
  ggplot(aes(Care_Type, value, fill=variable)) +
  geom_bar(stat='identity', position=position_dodge()) +
  xlab("") +
  ylab("% of Australian children") +
  scale_fill_manual(values=c("steelblue","grey","gold")) +
  theme_minimal() +
  theme(legend.title = element_blank(), legend.position="bottom")
```

##Childcare and age

Patterns of childcare do tend to change as children age. (Few Australian one-year-olds, for example, tend to attend before and/or after school care). Generally, younger children tend to be more likely to attend informal care, and older children more formal types of care. Below, for example, is the care profile for 2017.

```{r, echo=FALSE, warning=FALSE}
master %>%
  select(Care_Type, `Under 1_prop_2017`:`9–12_prop_2017`, `Under 1_prop_2014`: `9–12_prop_2014`, `Under 1_prop_2011`:`9–12_prop_2011`) %>%
  filter(!grepl(paste(filterwords, collapse="|"), Care_Type)) %>%
  filter(Care_Type!="Occasional care") %>%
  rename_at(vars(-Care_Type), function(x) gsub("_prop","", x)) %>%
  reshape2::melt(id.vars="Care_Type") %>%
  mutate(Year = case_when(
    str_detect(variable, "_2017")  ~ "2017",
    str_detect(variable, "_2014") ~ "2014",
    TRUE ~ "2011"
  )) %>%
  mutate(variable = gsub("_.*","",variable),
         variable = gsub("Under 1", "0", variable)) %>%
  ggplot(aes(variable, value, color=Year, group=Year)) +
  geom_line() +
  facet_wrap(~Care_Type) +
  ylab("% Australian children") +
  xlab("Age in years") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90)) +
  theme(legend.title = element_blank())
```

Long day care peaks for students at the age of 3-4, whilst grandparent care, as a proportion of all care, peaks when children are around 1. 

##Hours of care

However, the vast majority of grandparent care occurs for less than 10 hours a week. Let's check this out.

```{r, echo=FALSE}
master %>%
  select(Care_Type, `Less than 5_prop_2017`:`45 or more_prop_2017`) %>%
  filter(!grepl(paste(filterwords, collapse="|"), Care_Type)) %>%
  filter(Care_Type!="Occasional care") %>%
  rename_at(vars(-Care_Type), function(x) gsub("_prop_2017","", x)) %>%
  reshape2::melt(id.vars="Care_Type") %>%
  mutate(variable = gsub("Less than 5","0_5", variable),
         variable = gsub("45 or more","45+", variable),
         variable = gsub("–","_",variable)) %>%
  mutate(variable = factor(variable, levels=c("0_5","5_9","10_19","20_29","30_34","35_39","40_44","45+"))) %>%
ggplot(aes(variable, value, group=Care_Type, color=Care_Type)) +
  geom_line() +
  facet_wrap(~Care_Type) +
  theme(legend.title = element_blank(), axis.text.x = element_text(angle=90)) +
  xlab("Hours of care") +
  ylab("% of Australian children") +
  theme(legend.position = "none")
```

```{r, echo=FALSE}
master %>%
  select(Care_Type, `Less than 5_no_2017`:`45 or more_no_2017`) %>%
  mutate(Total = rowSums(.[2:9])) %>%
  kable()
  
```


















